---
title: "DEG Analysis"
author: "Ramya"
date: "2025-03-25"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


##1.Setup working directory

```{r }
setwd("C:/Users/Student/Downloads/Count_Practice/Count_Practice")

```


##2.Load Count data

```{r}
# Load library
library(gtools)

# Create a sorted list of text files
filelist <- mixedsort(list.files(pattern = "*.txt"))

# Read the count data from the file list
data_list <- lapply(filelist, function(fn) read.delim(fn, header = FALSE, sep = "\t")[,2]) 

# Combine data into a matrix
countdata <- do.call(cbind, data_list)

# Use the first column of a sample file as row names
rownames(countdata) <- read.delim(filelist[1], header = FALSE, sep = "\t")[,1] 
```


##3.Polish count data

```{r}
# Assign column names based on file names
colnames(countdata) <- sub(".count.txt", "", filelist)

# Convert data to numeric 
mode(countdata) <- "numeric"   

# Remove the last 5 rows showing summary 
countdata <- countdata[-c((nrow(countdata)-4):nrow(countdata)),]  

# View the data
head(countdata)
tail(countdata)
View(countdata)
```


##4.Generate colData

```{r}
# Extract sample names
sample.names <- colnames(countdata) 

# Create coldata table
coldata <- data.frame(
  name = sample.names,
  condition = rep(c("control", "treated"), each = 2, times = 2),
  sample = rep(c("A", "B"), each = 4)
)  

# Convert to factors (required by DESeq2)
coldata$condition <- factor(coldata$condition)  
coldata$sample <- factor(coldata$sample) 

 # View the coldata table
coldata   

# Save the table
write.csv(coldata, file="coldata.csv", row.names = FALSE)  
```


##5.Prepare DESeq2 data

```{r}
# Load library
library(DESeq2)

# Make a data frame form a matrix
dds <- DESeqDataSetFromMatrix(
  countData = countdata,
  colData = coldata,
  design = ~ condition + sample
)
```


##6.Plot dispersion of genes

```{r}
# Run basic DESeq2
dds <- DESeq(dds)

# Dispersion plot
plotDispEsts(dds, main = "Dispersion plot")
```


##7.Activity

```{r}
plotDispEsts(dds, main = "Dispersion Plot",
             genecol = "yellow",
             fitcol = "orange",
             finalcol = "pink",
             legend = TRUE,
             xlab = "Normalized counts",
             ylab = "Dispersion level",
             log = "xy",
             cex = 0.2)
```


##8.Normalization using DESeq2

```{r}
# Check the size factors calculated by geometric mean
dds.SF <- estimateSizeFactors(dds)
print(sizeFactors(dds.SF))

# Extract the normalized counts as transcript levels
Normalizedcount <- counts(dds.SF, normalized = T)
head(Normalizedcount)

# Save the normalized counts
write.csv(Normalizedcount, file="Normalized_counts.csv", row.names = TRUE)
```


##9.Plot counts

```{r}
# Assign colors for conditions
group.col <- c("skyblue", "orange")[coldata$condition]

#Plot the log2-transformed counts before and after normalization
png(filename = "../plot_counts.png", width = 6, height = 6, units = "in", res = 300)
par(mfrow=c(1,2), mar = c(1, 2, 1, 1))
boxplot(log2(counts(dds.SF, normalized=FALSE)+1),
        xlab="log2(counts+1)", main="Raw counts", las = 1,
        horizontal = TRUE, pch = ",", col = group.col)
boxplot(log2(counts(dds.SF, normalized=TRUE)+1),
        xlab="log2(normalized counts+1)", main="Normalized counts", las = 1,
        horizontal = TRUE, pch = ",", col = group.col)

# Close the graphics device
dev.off()
```


##10.Sample distance examination

```{r}
# Load libraries
library(gplots)
library(ggplot2)

# Variance-Stabilizing Transformation (VST)
transformed_data <- vst(dds, blind = TRUE)

# Calculate sample distance based on transformed data
sampleDists <- dist( t( assay(transformed_data)))

# Convert the distances into a matrix
sampleDistMatrix <- as.matrix(sampleDists)
```


##11.Heatmap

```{r}
# Use a color palette.
colours <- colorRampPalette(c("orange", "white"))(100)

# Create a heatmap for the sample distance
heatmap(
  sampleDistMatrix,
  Rowv = NA,   # Disable row dendrogram  
  Colv = NA,   # Disable column dendrogram  
  col = colours,  # Apply custom colors
  margins = c(8, 8),  # Adjust margins  
  main = "Sample distance"
)
```


##12.Principal Component Analysis (PCA)

```{r}
# Generate PCA data
pcaData <- plotPCA(transformed_data, intgroup = c("sample", "condition"), returnData = TRUE)

# Retrieve the percentage of variance explained by PC1 and PC2
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Create PCA Plot with ggplot2
PCA <- ggplot(pcaData, aes(PC1, PC2, color = condition, shape = sample)) +
  geom_point(size = 3, alpha = 0.8) +
  xlab(paste0("PC1 (", percentVar[1], "%) ")) +
  ylab(paste0("PC2 (", percentVar[2], "%) ")) +
  theme_bw() +
  ggtitle("PCA PLOT") + 
  theme(plot.title = element_text(size = 9, face = "bold"),
    legend.position = "top",
    axis.title = element_text(size = 17),   
    axis.text = element_text(size = 12),    
    panel.grid.major = element_blank(),     
    panel.grid.minor = element_blank() + 
    scale_color_manual(values = c("skyblue", "orange"))
  )

# view the plot
PCA

# Save the PCA plot as a PNG file using ggsave.
ggsave("../PCA_plot.png", plot = PCA, width = 5, height = 4, dpi = 300)
```


##13.Filtering

```{r}
# Create a copy of the dataset
ddsFiltered <- dds

# Define filtering criteria (only genes with sufficient expression (normalized counts ≥ 10) in at least 2 samples are retained.)
filterGenes <- rowSums(counts(ddsFiltered, normalized = TRUE)>= 10)>= 2

# Apply the filter to retain only genes meeting the criteria
ddsFiltered <- ddsFiltered[filterGenes, ]

# Plot dispersion after filtering
plotDispEsts(ddsFiltered, main="Dispersion after filtering")
```


##14.Run DESeq function

```{r}
# Update Design Formula including interaction terms between sample and condition.
design(ddsFiltered) <- ~sample + condition + sample:condition

# Run DESeq2 for normalization, dispersion estimation, and model fitting.
ddsFiltered <- DESeq(ddsFiltered)

# Check the DESeq2 comparisons
resultsNames(ddsFiltered)

# Extract Results for Sample A
resSampleA <- results(ddsFiltered, name = "condition_treated_vs_control")
summary(resSampleA)

# Extract Results for Sample B
resSampleB <- results(ddsFiltered, name = "sampleB.conditiontreated")
summary(resSampleB)
```


##15.Extract significant DEGs

```{r}
# Filter significant DEGs for Sample A
sigDEGs_A <- subset(resSampleA, padj<0.05 & abs(log2FoldChange) > 1)
sigDEGs_A
write.csv(as.data.frame(sigDEGs_A), file = "SampleA_DEGs.csv")

# Filter significant DEGs for Sample B
sigDEGs_B <- subset(resSampleB, padj<0.05 & abs(log2FoldChange) > 1)
sigDEGs_B
write.csv(as.data.frame(sigDEGs_B), file = "SampleB_DEGs.csv")
```


##16.Volcano plot

```{r}
library(ggplot2)

# Remove rows with NA in adjusted P values
resSampleA <- resSampleA[!is.na(resSampleA$padj), ]

# Prepare data for Sample A
resSampleA$Significance <- ifelse(resSampleA$padj < 0.05 & abs(resSampleA$log2FoldChange) > 1, "Significant", "Not Significant")

# Create the plot after loading ggplot2
volcanoA <- ggplot(resSampleA, aes(x = log2FoldChange, y = -log10(padj), color = Significance)) +  
geom_point(alpha = 0.8, size = 1.5) +  
scale_color_manual(values = c("Significant" = "red", "Not Significant" = "gray")) +  
labs(title = "Sample A",       
x = "Log2 Fold Change",      
y = "-Log10 Adjusted P-value")

volcanoA
```
###17. Activity
```{r}
library(ggplot2)

# Remove rows with NA in adjusted P values
resSampleA <- resSampleA[!is.na(resSampleA$padj), ]

# Categorize DEGs into Upregulated, Downregulated, and Not Significant
resSampleA$Significance <- ifelse(resSampleA$padj < 0.05 & resSampleA$log2FoldChange > 1, "Upregulated",
                            ifelse(resSampleA$padj < 0.05 & resSampleA$log2FoldChange < -1, "Downregulated", 
                                   "Not Significant"))

# Create the plot after loading ggplot2
volcanoA <- ggplot(resSampleA, aes(x = log2FoldChange, y = -log10(padj), color = Significance)) +  
  geom_point(alpha = 0.8, size = 1.8, shape = 16) +
  scale_color_manual(values = c("Upregulated" = "red", 
                                "Downregulated" = "blue", 
                                "Not Significant" = "gray")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.4) + 
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", linewidth = 0.4) +
  labs(title = "Volcano Plot: Sample A",
       x = expression(Log[2]~Fold~Change),
       y = expression(-Log[10]~"Adjusted P-value"),
       color = "Gene Regulation") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank()
  )

volcanoA
ggsave("volcano_sampleA_custom.png", volcanoA, width = 8, height = 6, dpi = 300)
```
###18. Sample B
```{r}
library(ggplot2)

# Remove rows with NA in adjusted P values
resSampleB <- resSampleB[!is.na(resSampleB$padj), ]

# Categorize DEGs into Upregulated, Downregulated, and Not Significant
resSampleB$Significance <- ifelse(resSampleB$padj < 0.05 & resSampleB$log2FoldChange > 1, "Upregulated",
                            ifelse(resSampleB$padj < 0.05 & resSampleB$log2FoldChange < -1, "Downregulated", 
                                   "Not Significant"))

# Create the plot after loading ggplot2
volcanoB <- ggplot(resSampleB, aes(x = log2FoldChange, y = -log10(padj), color = Significance)) +  
  geom_point(alpha = 0.8, size = 1.8, shape = 16) +
  scale_color_manual(values = c("Upregulated" = "orange", 
                                "Downregulated" = "red", 
                                "Not Significant" = "gray")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.4) + 
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", linewidth = 0.4) +
  labs(title = "Volcano Plot: Sample B",
       x = expression(Log[2]~Fold~Change),
       y = expression(-Log[10]~"Adjusted P-value"),
       color = "Gene Regulation") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    panel.grid.major = element_line(color = "gray"),
    panel.grid.minor = element_blank()
  )

volcanoB
ggsave("volcano_sampleB_custom.png", volcanoB, width = 8, height = 6, dpi = 300)
```


#19.Plot Specific Genes
###Prepare normalized count data with coldata

```{r}
library(Rmisc)

# Read and process normalized counts
NormalizedCount <- read.csv("Normalized_counts.csv", row.names = 1)
NC_t <- cbind(name = rownames(t(NormalizedCount)), t(NormalizedCount))

# Load coldata and restore 'name' column
coldata <- read.csv("coldata.csv", header = TRUE, row.names = 1)
coldata$name <- rownames(coldata)

# Merge coldata with normalized counts
NC_col <- merge(coldata, NC_t, by = "name")

# Ensure all columns from the 4th to the last are numeric
NC_col[, 4:ncol(NC_col)] <- lapply(NC_col[, 4:ncol(NC_col)], function(x) as.numeric(as.character(x)))
```


###20.Summarize data for a gene
#Plotting

```{r}
# Summarize data for an example gene, Glyma_13G122766
Glyma_13G122766.sum <- summarySE(NC_col, measurevar = "Glyma_13G122766", groupvars = c("sample", "condition"))

# Check the summary
Glyma_13G122766.sum
```

```{r}
# Create the plot for Glyma_13G122766
Glyma_13G122766 <- ggplot(Glyma_13G122766.sum, aes(x = sample, y = Glyma_13G122766, fill = condition)) +
geom_bar(position = position_dodge(), stat = "identity", color = "grey20") +
  labs(y = "Transcript levels\n(normalized counts)", 
       x = "Sample", 
       title = "Glyma_13G122766") +
  geom_errorbar(aes(ymin = Glyma_13G122766, ymax = Glyma_13G122766 + sd),
 linewidth = 0.3, color = "grey20", width = 0.2, position = position_dodge(0.9)) +
  scale_fill_manual(values = c("skyblue", "orange"),
                    breaks = c("control", "treated"))

# check the plot
Glyma_13G122766

# Save the plot as a PNG file
ggsave(filename = "Glyma_13G122766.png",
       plot = Glyma_13G122766, 
       width = 4, height = 4, units = "in", dpi = 300)
```

###21.Activity

```{r}
# Summarize data for an example gene, Glyma_15G074200
Glyma_15G074200.sum <- summarySE(NC_col, measurevar = "Glyma_15G074200", groupvars = c("sample", "condition"))

# Check the summary
Glyma_15G074200.sum

# Create the plot for Glyma_15G074200
Glyma_15G074200.plot <- ggplot(Glyma_15G074200.sum, aes(x = sample, y = Glyma_15G074200, fill = condition)) +
  geom_bar(position = position_dodge(), stat = "identity", color = "black", width = 0.7) +
  geom_errorbar(aes(ymin = Glyma_15G074200 - sd, ymax = Glyma_15G074200 + sd),
                width = 0.2, color = "black", position = position_dodge(0.7)) +
  labs(title = "Expression of Glyma_15G074200",
       x = "Sample",
       y = "Transcript levels") +
  scale_fill_manual(values = c("control" = "green", "treated" = "red"),
                    name = "Condition",
                    labels = c("Control", "Infested")) +
   theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "gray"),
    panel.grid.minor = element_blank()
  )

# check the plot
Glyma_15G074200.plot

# Save the plot as a PNG file
ggsave(filename = "Glyma_15G074200_expression_plot.png",
       plot = Glyma_15G074200.plot,
       width = 5, height = 4.5, units = "in", dpi = 300)
```






































































